XRONOS: an open data infrastructure for archaeological chronology

### 2024-10-13 paper_jcaa.qmd
Expand draft

```{r setup}
library("countrycode")
library("cowplot")
library("dplyr", warn.conflicts = FALSE)
library("giscoR")
library("ggplot2")
library("glue")
library("khroma")
library("patchwork", warn.conflicts = FALSE)
library("readr")
library("sf")
library("tidyr")
library("tinytable")

@@

<!--
* [ ] First draft:
  * [ ] Introduction [500]
  * [ ] State of the art [2000]
    * [ ] Compilations of radiocarbon dates [1000]
    * [x] Open data infrastructures [500]
    * [ ] Beyond radiocarbon [500]
  * [ ] Concept [1000]
    * [x] Inherits a common structure from RADON
    * [ ] Design goals [250]
    * [x] Data model [1150/500]
    * [x] LOD [250/250]
  * [x] Implementation [1000]
    * [x] Status of development [100/100]
    * [x] Software architecture [550/300]
    * [x] Data ingestion and integration [500/300]
    * [x] User interfaces [650/300]
  * [ ] Evaluation [500]
* [ ] Add references:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
* [ ] Figures and tables
  * [x] FIGURE: graph of number of datasets through time
  * [x] FIGURE: map of number of dates in datasets
  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
  * [ ] FIGURE: data model
  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
  * [ ] TABLE: list of 'issues'
  * [ ] APPENDIX: table of datasets
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

```{r fig-c14-datasets-map}
#| fig-cap: Geographic coverage of published regional radiocarbon compilations
c14_datasets |>
  separate_longer_delim(m49_region, "; ") |>
  drop_na(m49_region) |>
  summarise(n_dates = sum(n_dates, na.rm = TRUE), .by = m49_region) |>
  right_join(m49_regions) |>
  mutate(n_dates = replace_na(n_dates, 0)) |>
  st_as_sf() |>
  ggplot() +
  geom_sf(aes(fill = n_dates), colour = NA) +
  geom_sf(data = coast, fill = NA) +
  scale_fill_turku(reverse = TRUE, labels = scales::comma) +
  labs(fill = "Dates") +
  coord_sf(crs = "+proj=moll") +
  theme_map()
```

Although this body of work has greatly improved the accessibility of radiocarbon dates and supported significant methodological advances [@CremaBevan2021], some limitations are apparent.

@@

Compilations also become out of date almost immediately upon publication, due the the constant stream of new dates.
Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` had received updates in the last two years.
The average 'lifespan' of a dataset from its publication to its last update is `r mean_c14_datasets_lifespan` years.
Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g. a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.

@@

@@

- The first available synthetic radiocarbon database was **c14bazAAR** [@CITE], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (see @FIG).

- Another indexical approach is the **IntChron** project [@CITE], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common itnerface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
The JSON format also lends itself to the implementation of wrapper libraries, for example the 'rIntChron' package gives direct access to IntChron-indexed databases in R [@CITE].

* **p3k14c** [@CITE] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR. The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses. The downside is that without the continuous link to the source databases present in the c14bazAAR and IntChron, it can only be kept up to date manually with periodic re-releases. An accompanying package (@CITE) provides direct access to the p3k14c dataset in R.

@@

<!-- TODO: probably should be in design goals or evaluation -->
XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
However in general we see the approaches as complementary.
A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
New data and corrections from p3k14c are incorporated into XRONOS as they are released.

@@

@@

XRONOS inherits its basic structure from RADON (@CITE), with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for XX years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.

@@

<!-- TODO: Look at Martin's grant application and drafts -->

@@

Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model (@FIG?), and we use thesauruses to automatically standardise these fields as much as possible.
For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy (@CITE), and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
A wide variety of other potential data quality issues (e.g. missing data on what country a site is in) are also flagged for human review by this system, which canoften be semi-automated (e.g. suggesting close matches in the thesaurus or the country indicated by the record's coordinates).

@@

@@

>>>>

```{r setup}
library("countrycode")
library("cowplot")
library("dplyr", warn.conflicts = FALSE)
library("giscoR")
library("ggplot2")
library("glue")
library("gt")
library("khroma")
library("patchwork", warn.conflicts = FALSE)
library("readr")
library("sf")
library("tidyr")

@@

<!--
* [ ] First draft:
  * [ ] Introduction [500]
  * [ ] State of the art [2000]
    * [x] Compilations of radiocarbon dates [1000]
    * [x] Open data infrastructures [500]
    * [ ] Beyond radiocarbon [500]
  * [ ] Concept [1000]
    * [x] Inherits a common structure from RADON
    * [ ] Design goals [250]
    * [x] Data model [1150/500]
    * [x] LOD [250/250]
  * [x] Implementation [1000]
    * [x] Status of development [100/100]
    * [x] Software architecture [550/300]
    * [x] Data ingestion and integration [500/300]
    * [x] User interfaces [650/300]
  * [ ] Evaluation [500]
* [ ] Add references:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  * [ ] Appendix table
* [ ] Figures and tables
  * [x] FIGURE: graph of number of datasets through time
  * [x] FIGURE: map of number of dates in datasets
  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
  * [ ] FIGURE: data model
  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
  * [ ] TABLE: list of 'issues'
  * [ ] APPENDIX: table of datasets
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

```{r fig-c14-datasets-map}
#| fig-cap: Geographic coverage of published regional radiocarbon compilations
c14_datasets |>
  separate_longer_delim(m49_region, "; ") |>
  drop_na(m49_region) |>
  summarise(n_dates = sum(n_dates, na.rm = TRUE), .by = m49_region) |>
  right_join(m49_regions) |>
  mutate(n_dates = replace_na(n_dates, 0)) |>
  st_as_sf() |>
  ggplot() +
  geom_sf(aes(fill = n_dates), colour = NA) +
  geom_sf(data = coast, fill = NA) +
  scale_fill_turku(reverse = TRUE, labels = scales::comma) +
  labs(fill = "Dates") +
  coord_sf(crs = "+proj=moll") +
  theme_minimal_grid()
```

Although this body of work has greatly improved the accessibility of radiocarbon dates and supported significant methodological advances [@Crema2022; @CremaEtAl2024], some limitations are apparent.

@@

Datasets based on literature review also become out of date almost immediately upon publication, due the the constant stream of new dates.
Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` had received updates in the last two years.
The average 'lifespan' of a dataset from its publication to its last update is `r mean_c14_datasets_lifespan` years.
Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g. a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.

<!-- TODO: lab databases don't go out of date, but have arbitary coverage? -->

* The coverage of these databases is uneven, systematically biased, and duplicative of each other [@CITE Africa paper]?

@@

<!-- TODO: -->
* Not all are open
* Not all are machine readable

@@

- The first available synthetic radiocarbon database was **c14bazAAR** [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (@FIG).

- Another indexical approach is the **IntChron** project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common itnerface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
The JSON format also lends itself to the implementation of wrapper libraries, for example the 'rIntChron' package gives direct access to IntChron-indexed databases in R [@CITE].

* **p3k14c** [@BirdEtAl2022] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR. The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses. The downside is that without the continuous link to the source databases present in the c14bazAAR and IntChron, it can only be kept up to date manually with periodic re-releases. An accompanying package (@CITE) provides direct access to the p3k14c dataset in R.

@@

@@

* There are lots of other types of chronometric data!
  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
  * Other radiometric/absolute dating methods, OSL, etc.
  * Typology - still the backbone of many chronologies
  * Application-specific: sea level dating, rock patina, etc.
* Compilation of other types of dates has been much more limited.
  * Exceptions: ...?

<!-- MH to do: -->
For European dendrochronological data, the situation is even less favourable for archaeological purposes:

* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
* DendroDB has a European focus but only provides data for historical periods.
* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD). An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g. ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%). Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).

* Typology
  * perio.do
  * Often embedded in other types of database (e.g. gazetteers)
  * But nobody has treated it as a form of chronometric data in its own right?

What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data. However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.

@@

XRONOS inherits its basic structure from RADON [@Raetzel-Fabian1999; @RADON; @RADONB; @RADONB2024], with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for XX years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.

@@

The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
develop robust tools for the continuous ingestion and refinement of this data; 
and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.

Our concrete goals in the design and implementation of XRONOS are the following:

1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
4. licensing and versioning of individual data using eg. DOI
5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies

@@

```{r tbl-issues}
#| tbl-cap: Automatically-recognised data quality issues currently implemented in XRONOS
tribble(
  ~entity,                    ~issue_code,            ~n,     ~issue_description,
  "Sites",                    "MISSING_COORDINATES",  4452,   "Missing geographic coordinates",
  "Sites",                    "INVALID_COORDINATES",  0,      "Geographic coordinates fall outside the earth's ellipsoid",
  "Sites",                    "MISSING_COUNTRY_CODE", 1221,   "Missing data on what country the site is in",
  "Samples",                  "MISSING_MATERIAL",     138177, "Missing data on the sample material",
  "Samples",                  "MISSING_TAXON",        138182, "Missing data on the sample taxon",
  "Samples",                  "MISSING_CRS",          0,      "Sample has coordinates but the coordinate reference system is not given",
  "Taxons",                   "UNKNOWN_TAXON",        9260,   "Sample taxon has not been matched to the GBIF Backbone Taxonomy",
  "Taxons",                   "LONG_TAXON",           509,    "Description of the sample taxon is implausibly long",
  "Radiocarbon dates",        "MISSING_C14_AGE",      447,    "Missing radiocarbon age",
  "Radiocarbon dates",        "VERY_OLD_C14",         76,     "Radiocarbon age older than the effective range of the method (50 ka)",
  "Radiocarbon dates",        "MISSING_C14_ERROR",    585,    "Missing measurement error",
  "Radiocarbon dates",        "MISSING_D14C",         238875, "Missing δ13C measurement",
  "Radiocarbon dates",        "MISSING_D14C_ERROR",   238875, "Missing δ13C measurement error",
  "Radiocarbon dates",        "MISSING_C14_METHOD",   242233, "Missing data on radiocarbon dating method (conventional, AMS, etc.)",
  "Radiocarbon dates",        "MISSING_C14_LAB_ID",   1233,   "Missing laboratory identifier",
  "Radiocarbon dates",        "INVALID_LAB_ID",       16053,  "Laboratory identifier does not match the standard format (e.g. 'Abc-1234')",
  "Radiocarbon dates",        "MISSING_C14_LAB",      350190, "Missing data on the radiocarbon laboratory",
  "Bibliographic references", "MIXED_REFERENCE",      21933,  "Bibliographic reference appears to combine multiple publications",
  "Bibliographic references", "MISSING_BIBTEX",       42109,  "Bibliographic reference without structured data in BibTeX format",
) |>
  group_by(entity) |>
  gt() |>
  cols_label(issue_code = "Issue", n = "N", issue_description = "Description") |>
  cols_width(everything() ~ pct(33)) |> # TODO: fix...
  tab_options(
    table.font.size = 13
  )
```

Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model (@FIG?), and we use thesauruses to automatically standardise these fields as much as possible.
For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy (@CITE), and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
A wide variety of other potential data quality issues (e.g. missing data on what country a site is in) are also flagged for human review by this system [@tbl-issues], which can often be semi-automated (e.g. suggesting close matches in the thesaurus or the country indicated by the record's coordinates).

@@

This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data; 
an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
and an R package for interfacing with the API. 
We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.

@@

XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
However in general we see the approaches as complementary.
A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
New data and corrections from p3k14c are incorporated into XRONOS as they are released.

# Acknowledgements

The XRONOS project was carried under the direction of Albert Hafner. 
Individual contributors to the XRONOS database to date include Chiara Huwiler, Rivana Moser, Tomasz Chmielewski, and Stephanie Döppler. 
A complete and up-to-date list of acknowledgements for the project can be found at <https://xronos.ch/about/acknowledgements>.

# Funding statement

This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality').
